---
title: "Model Building"
author: "Joy Zhang"
date: "2023-12-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(ResourceSelection)
library(car)
library(corrplot)
library(dplyr)
library(Metrics)

diabetes_data <- read.csv("diabetes_data.csv")
```

# Factor covariates as necessary
```{r}
diabetes_data$Diabetes_binary <- factor(diabetes_data$Diabetes_binary)
diabetes_data$HighBP <- factor(diabetes_data$HighBP)
diabetes_data$HighChol <- factor(diabetes_data$HighChol)
diabetes_data$CholCheck <- factor(diabetes_data$CholCheck)
diabetes_data$Smoker <- factor(diabetes_data$Smoker)
diabetes_data$Stroke <- factor(diabetes_data$Stroke)
diabetes_data$HeartDiseaseorAttack <- factor(diabetes_data$HeartDiseaseorAttack)
diabetes_data$PhysActivity <- factor(diabetes_data$PhysActivity)
diabetes_data$Fruits <- factor(diabetes_data$Fruits)
diabetes_data$Veggies <- factor(diabetes_data$Veggies)
diabetes_data$HvyAlcoholConsump <- factor(diabetes_data$HvyAlcoholConsump)
diabetes_data$AnyHealthcare <- factor(diabetes_data$AnyHealthcare)
diabetes_data$NoDocbcCost <- factor(diabetes_data$NoDocbcCost)
diabetes_data$DiffWalk <- factor(diabetes_data$DiffWalk)
diabetes_data$Sex <- factor(diabetes_data$Sex)
diabetes_data$age_cat <- factor(diabetes_data$age_cat)
diabetes_data$income_cat <- factor(diabetes_data$income_cat)
diabetes_data$edu_cat <- factor(diabetes_data$edu_cat)
diabetes_data$PhysHlth_cat <- factor(diabetes_data$PhysHlth_cat)
diabetes_data$MentHlth_cat <- factor(diabetes_data$MentHlth_cat)
```

# Fit first model with just age
```{r}
model1 <- glm(data = diabetes_data, formula = Diabetes_binary ~ age_cat, family = binomial("logit"))
summary(model1)
```

# Fit model with all covariates of interest (No BMI, no GenHealth)
```{r}
model2 <- glm(data = diabetes_data, formula = Diabetes_binary ~ age_cat + HighBP + HighChol + CholCheck + Smoker + Stroke + HeartDiseaseorAttack + PhysActivity + Fruits + Veggies + HvyAlcoholConsump + AnyHealthcare + NoDocbcCost + DiffWalk + Sex + income_cat + edu_cat + PhysHlth_cat + MentHlth_cat, family = binomial("logit"))
summary(model2)
library(xtable)
xtable(model2)
```

# Compare models to see which is better
```{r}
anova(model1, model2, test = "LRT")
```
Significant p-value from LRT says model 2 (adjusted model) is better - we proceed with model 2 to run diagnostic tests. 

# Check for multicollinearity using VIF
```{r}
model2_vif <- vif(model2)
vif_res <- t(data.frame(VIF = model2_vif))
kable(vif_res, digits=2)
```
Most values are near 1, suggesting low multicollinearity

# Goodness of Fit using Pearson's Chi-square test
```{r}
# Pearson Chi-square Goodness of Fit Test
PChiAgeLog <- sum(residuals(model2, type = "pearson") ^ 2)
df <- nrow(diabetes_data) - 20
p_value <- pchisq(PChiAgeLog, df, lower.tail = FALSE)

cat("Chi-square statistic:", PChiAgeLog, "\n")
cat("degrees of freedom:", df, "\n")
cat("p-value:", p_value, "\n")
```
P-value is insignificant, suggests model is a good fit

# Plotting fitted vs. Resid
```{r}
plot(x = fitted.values(model2), y = residuals(model2, type = "pearson"),
     col = ifelse (diabetes_data[, "Diabetes_binary"] == 1, "darkblue", "chartreuse4"))
```

# Fisher test
```{r, include=FALSE}
Fisher <- matrix(nrow = ncol(diabetes_data), ncol = ncol(diabetes_data))

for (i in 1:ncol(diabetes_data)) {
  for (j in 1:ncol(diabetes_data)) {
    # only perform test for unique pairs
    if (i < j) {
      # perform Fisher's exact test
      result <- fisher.test(as.data.frame(diabetes_data)[,i], as.data.frame(diabetes_data)[,j],
                            simulate.p.value = TRUE)
      # store p-value in matrix
      Fisher[i, j] <- result$p.value
    }
  }
}

colnames(Fisher) <- colnames(diabetes_data)

Fisher 

significant_pairs <- which(Fisher < 0.001, arr.ind = TRUE)

# Print the variable pairs with significant correlation
for (row in 1:nrow(significant_pairs)) {
  i <- significant_pairs[row, 1]
  j <- significant_pairs[row, 2]
  cat("Significant correlation:", colnames(Fisher)[i], "*", colnames(Fisher)[j], "\n")
}
```

# Deviance
```{r}
residual_deviance <- model2$deviance
df <- model2$df.residual

ratio_deviance_df <- residual_deviance / df

print(ratio_deviance_df)
```

# Cook's Distance
```{r}
influence_info <- influence.measures(model2)
cooksd <- influence_info$infmat[, "cook.d"]
leverage <- influence_info$infmat[, "hat"]
std_resid <- influence_info$infmat[, ncol(influence_info$infmat)]

plot(cooksd, pch = 20, main = "Cook's Distance Plot", ylab = "Cook's Distance")

influential_points <- which(cooksd > 4/nrow(influence_info$infmat))
points(influential_points, cooksd[influential_points], col = "red", pch = 20)

plot(leverage, pch = 20, main = "Leverage Plot", ylab = "Leverage")

high_leverage_points <- which(leverage > 2 * mean(leverage))
points(high_leverage_points, leverage[high_leverage_points], col = "red", pch = 20)
```

